---
title: "otter_census_analysis"
author: "Owen Liu"
date: "March 14, 2016"
output: html_document
---

```{r setup, echo = FALSE, message = FALSE}
library(plyr)
library(dplyr)
library(tidyr)
library(rgdal)
library(stringr)
library(maptools)
library(ggplot2)
library(sp)
library(rgeos)
library(broom)
library(reshape)
library(reshape2)
```


Starting with the 1985-1994 census data
```{r read shapefiles 1, echo = TRUE, eval = TRUE}
dir_spatial   <- '~/github/sea_otter_census/data/seaottercensusdata1985_1994/sp85' 
layer_85 <- 'Census_sum_85'

poly_census_85 <- readOGR(dsn = path.expand(dir_spatial), layer = layer_85, stringsAsFactors = FALSE)

# poly_census_85.tidy <- tidy(poly_census_85,region='POLY_ID')
# ggplot(poly_census_85.tidy, aes(long,lat)) +geom_polygon()
```


```{r viewing SpatialPolygonsDataFrame slots, echo = TRUE, eval = TRUE}
### view the overall structure
summary(poly_census_85)

### check out the attribute table in the data slot
head(poly_census_85@data)

### check out the CRS info
poly_census_85@proj4string
```

Plot ATOS by Density
```{r density by latitude, echo=TRUE, eval=TRUE}
## Data is stored in the @data slot of the SpatilPolygonsDataFrame
atts_85<-poly_census_85@data
names(atts_85)
dens_85<-ggplot(atts_85,aes(x=ATOS_ID,y=lin_dens)) + geom_smooth() +xlab('Census section (North to South)') +ylab('Density per km coastline') +ggtitle('Otter Density by Latitude')
dens_85
```

Import all the other years and see if we can do the same analyses
```{r importing all shapefiles, echo=FALSE}

### for 1985 to 1994

atts_all <- list()

for(i in 85:94) {
  dir_spatial  <- paste0('~/github/sea_otter_census/data/seaottercensusdata1985_1994/sp',i)
  layer <- paste0('Census_sum_',i)

  poly_yr <- readOGR(dsn = path.expand(dir_spatial), layer = layer, stringsAsFactors = FALSE)
  poly_name <- paste0('poly_census_',i)
  assign(poly_name,poly_yr)
  
  atts<-poly_yr@data
  atts_name <- paste0('atts_',19,i)
  plot_name <- paste0('dens_',19,i)
  dens_plot<-ggplot(atts,aes(x=ATOS_ID,y=lin_dens)) + geom_smooth() +xlab('Census section (North to South)') +ylab('Density per km coastline') +ggtitle(paste0('Otter Density by Latitude, Year: ',i)) +ylim(c(0,5))
  assign(plot_name,dens_plot)
  assign(atts_name,atts)
  atts_all[[atts_name]] <- atts
}


### for 1995 to 2004
for(i in 95:99) {
  dir_spatial  <- paste0('~/github/sea_otter_census/data/seaottercensusdata1995_2004/sp',i)
  layer <- paste0('Census_sum_',i)

  poly_yr <- readOGR(dsn = path.expand(dir_spatial), layer = layer, stringsAsFactors = FALSE)
  poly_name <- paste0('poly_census_',i)
  assign(poly_name,poly_yr)
  
  atts<-poly_yr@data
  atts_name <- paste0('atts_',19,i)
  plot_name <- paste0('dens_',19,i)
  dens_plot<-ggplot(atts,aes(x=ATOS_ID,y=lin_dens)) + geom_smooth() +xlab('Census section (North to South)') +ylab('Density per km coastline') +ggtitle(paste0('Otter Density by Latitude, Year: ',i)) +ylim(c(0,5))
  assign(plot_name,dens_plot)
  assign(atts_name,atts)
  atts_all[[atts_name]] <- atts
}

### for 2000 to 2004
for(i in c('00','01','02','03','04')) {
  dir_spatial  <- paste0('~/github/sea_otter_census/data/seaottercensusdata1995_2004/sp',i)
  layer <- paste0('Census_sum_',i)

  poly_yr <- readOGR(dsn = path.expand(dir_spatial), layer = layer, stringsAsFactors = FALSE)
  poly_name <- paste0('poly_census_',i)
  assign(poly_name,poly_yr)
  
  atts<-poly_yr@data
  atts_name <- paste0('atts_',20,i)
  plot_name <- paste0('dens_',20,i)
  dens_plot<-ggplot(atts,aes(x=ATOS_ID,y=lin_dens)) + geom_smooth() +xlab('Census section (North to South)') +ylab('Density per km coastline') +ggtitle(paste0('Otter Density by Latitude, Year: ',i)) +ylim(c(0,5))
  assign(plot_name,dens_plot)
  assign(atts_name,atts)
  atts_all[[atts_name]] <- atts
}

### for 2005 to 2014
for(i in c('05','06','07','08','09',10,12,13,14)) {
  dir_spatial  <- paste0('~/github/sea_otter_census/data/seaottercensusdata2005_2014/sp',i)
  layer <- paste0('Census_sum_',i)

  poly_yr <- readOGR(dsn = path.expand(dir_spatial), layer = layer, stringsAsFactors = FALSE)
  poly_name <- paste0('poly_census_',i)
  assign(poly_name,poly_yr)
  
  atts<-poly_yr@data
  atts_name <- paste0('atts_',20,i)
  plot_name <- paste0('dens_',20,i)
  dens_plot<-ggplot(atts,aes(x=ATOS_ID,y=lin_dens)) + geom_smooth() +xlab('Census section (North to South)') +ylab('Density per km coastline') +ggtitle(paste0('Otter Density by Latitude, Year: ',i)) +ylim(c(0,5))
  assign(plot_name,dens_plot)
  assign(atts_name,atts)
  atts_all[[atts_name]] <- atts
}
```

Range extent over time

```{r range extent, echo=TRUE,eval=TRUE}
calc_range <- function(x) c((range(x$ATOS_ID)), (max(x$ATOS_ID)-min(x$ATOS_ID))/2)
ranges <- t(sapply(atts_all,calc_range))
range.out <- data.frame(year=c(1985:2010,2012:2014),min=ranges[,1],max=ranges[,2],range=ranges[,3])

range_plot<-ggplot(range.out) + geom_line(aes(x=year,y=range)) + xlab('Year')+ylab('Range Extent, km of coastline') +ggtitle('Total Range Extent by Year, 1985-2014')

northern_edge <- ggplot(range.out) + geom_path(aes(x=year,y=min)) + xlab('Year')+ylab('Northern Range Limit (most northerly ATOS point') +ggtitle('Northern Extent of Range by Year, 1985-2014') + scale_y_reverse()

southern_edge <- ggplot(range.out) + geom_path(aes(x=year,y=max)) + xlab('Year')+ylab('Southern Range Limit (most southerly ATOS point)') +ggtitle('Southern Extent of Range by Year, 1985-2014')

range_plot
northern_edge
southern_edge
# note: no data for 2011 because of weather, so range is average of 2013 and 2014

```


We are interested partly in how otter density has changed with marine protected area implementation.  We have a dataset for California MPAs, let's add it.

```{r California MPAs, echo=TRUE, eval=TRUE}
dir_spatial  <- paste0('~/github/sea_otter_census/data/MPA_CA_Existing_160301')
layer <- 'MPA_CA_Existing_160301'
cal_mpas <- readOGR(dsn = path.expand(dir_spatial), layer = layer, stringsAsFactors = FALSE)
cal_mpas.data <- cal_mpas@data

plot(cal_mpas)
# luckily, this layer has the same Proj4String as the otter data, so we don't have to worry as much about spatial congruence of the layers

### view the overall structure
summary(cal_mpas)

### check out the attribute table in the data slot
head(cal_mpas@data)

### spatial overlap between an otter layer and the marine reserves
### over() function returns, in a list, all the elements of the otter census that touch or overlap any of the mpas, then we extract only the overlapping IDs and use the unique overlapping POLY_IDs as our index of otter census tracts that are associated with marine reserves.

## list of all MPAs and which census tracts overlap with them
otter_mpa_over <- over(cal_mpas,poly_census_09,returnList = T)
## assign the MPA ID #s to the overlay list so we can properly reference them later
names(otter_mpa_over)<-cal_mpas@data$OBJECTID

## just a list of the overlaps
overlaps <- otter_mpa_over[sapply(otter_mpa_over, function(x) dim(x)[1]) > 0]

## subset (IDs) of MPAs that are within the otter range
over_mpas <- cal_mpas[sapply(otter_mpa_over, function(x) dim(x)[1]>0),]

## census polygon IDs that fall within overlapping MPAs
over_polyIDs<- sapply(overlaps,function(x) unique(x$POLY_ID))
#and as a simple vector
polyIDs.vec <-unlist(over_polyIDs)

#visual check-- census polygons that fall within or touch MPAs
census_inside <- poly_census_09[which(poly_census_09@data$POLY_ID %in% polyIDs.vec),]
plot(census_inside)

#visual check-- census polygons that DO NOT fall within or touch MPAs
census_outside <- poly_census_09[which(!poly_census_09@data$POLY_ID %in% polyIDs.vec),]
plot(census_outside)


```

So now we have our list of unique IDs.  Because we used the year of census data with the largest spatial extent (largest otter range, 2009), we should be able to apply those IDs too all of the years of data, and obtain densities for inside/outside MPAs for each year of census data.  We'll first write a function to do that for a given census year, then apply it to all years.  Of course, to do this we will also need the inverse selection, i.e. those census tracts outside the MPAs.

```{r census subset, echo=TRUE,eval=TRUE}

#keeper columns
keeps <- c('POLY_ID','lin_dens',"Year")
dens.all.years <- ldply(atts_all,function(x) x[keeps],.id=NULL)
dens.all.years$in_out <- apply(dens.all.years,MARGIN=1,function(x) ifelse(x["POLY_ID"] %in% polyIDs.vec, 'Inside',"Outside"))

# mean density by year
mpa.dens.by.yr <- aggregate(list(mean_dens=dens.all.years$lin_dens),by=list(Year=dens.all.years$Year,Reserve=dens.all.years$in_out),FUN=mean)


# standard deviation
mpa.dens.sd.by.yr <- aggregate(list(sd_dens=dens.all.years$lin_dens),by=list(Year=dens.all.years$Year,Reserve=dens.all.years$in_out),FUN=sd)
mpa.dens.by.yr <-merge(mpa.dens.by.yr,mpa.dens.sd.by.yr)

inside.by.yr <- subset(mpa.dens.by.yr,mpa.dens.by.yr$Reserve=='Inside')
outside.by.yr <- subset(mpa.dens.by.yr,mpa.dens.by.yr$Reserve=='Outside')

# time series plots of mean density inside and outside over time
ggplot(mpa.dens.by.yr,aes(x=Year,y=mean_dens,col=Reserve,group=Reserve)) + geom_line() +ylab('Otters/km of coastline') + scale_colour_manual("",labels = c("Inside MPAs", "Outside MPAs"),values = c("black", "blue")) +ggtitle('Otter Density Inside and Outside of MPAs')

# box plot of all density data inside/outsideR
ggplot(mpa.dens.by.yr,aes(x=Reserve,y=mean_dens)) + geom_boxplot()


### BACI
## Assumption is that marine reserves went into effect in 2007 (true for most central coast MPAs)
before <- subset(dens.all.years,dens.all.years$Year<=2007)
after <- subset(dens.all.years,dens.all.years$Year>2007)


dens.all.years$bef_aft <- apply(dens.all.years,MARGIN=1,function(x) ifelse(x["Year"] < 2008, 'Before',"After"))

bef.aft.in.out <- aggregate(list(mean_dens=dens.all.years$lin_dens),by=list(Reserve=dens.all.years$in_out, Period=dens.all.years$bef_aft),FUN=mean)

bef.aft.in.out$plotorder <- sapply(bef.aft.in.out$Period,function(x) ifelse(x=='Before','1','2'))


ggplot(bef.aft.in.out,aes(y=mean_dens,color=Reserve,group=Reserve)) + geom_line(aes(x=plotorder)) +ylab('Mean Density') +xlab('Period') +scale_x_discrete(labels=c('Before','After')) +scale_color_manual(values=c('seagreen','royalblue'))

## another column for just the poly number (without the character) for the next step
dens.all.years$polynum <- as.numeric(sapply(dens.all.years$POLY_ID,function(x) gsub("[[:alpha:]]","",x)))

```

Just a subset of MPAs (in this case, just the central coast)

```{r subset over time}
plot_density_subset <- function(poly_subset=c(400,800)) {
  dat.subset <- dens.all.years[dens.all.years$polynum >= poly_subset[1] & dens.all.years$polynum <=poly_subset[2],]
  dens.by.yr <- aggregate(list(mean_dens=dat.subset$lin_dens),by=list(Year=dat.subset$Year,Reserve=dat.subset$in_out),FUN=mean)
  # time series plots of mean density inside and outside over time
  dens.over.time <- ggplot(dens.by.yr,aes(x=Year,y=mean_dens,col=Reserve,group=Reserve)) + geom_line() +ylab('Otters/km of coastline') +   scale_colour_manual("",labels = c("Inside MPAs", "Outside MPAs"),values = c("black", "blue")) +ggtitle('Otter Density Inside and Outside of MPAs')
  
  # box plot of all density data inside/outside
  dens.boxplot<-ggplot(dens.by.yr,aes(x=Reserve,y=mean_dens)) + geom_boxplot()
  
  
  ### BACI
  bef.aft.in.out <- aggregate(list(mean_dens=dat.subset$lin_dens),by=list(Reserve=dat.subset$in_out, Period=dat.subset$bef_aft),FUN=mean)
  
  bef.aft.in.out$plotorder <- sapply(bef.aft.in.out$Period,function(x) ifelse(x=='Before','1','2'))
  
  baci.plot<-ggplot(bef.aft.in.out,aes(y=mean_dens,color=Reserve,group=Reserve)) + geom_line(aes(x=plotorder)) +ylab('Mean Density') +xlab('Period') +scale_x_discrete(labels=c('Before','After'))+scale_color_manual(values=c('seagreen','royalblue'))
  
  return(list(dens_over_time=dens.over.time,box=dens.boxplot,baci=baci.plot))
}

```

Specific to each MPA

```{r buffer and clip mpas, echo=TRUE, eval=TRUE}

#state layer for clipping to coast
dir_spatial  <- paste0('~/github/sea_otter_census/data/cnty24k09_1_shp/shapefiles')
layer <- 'cnty24k09_1_state_poly'
cal_land <- readOGR(dsn = path.expand(dir_spatial), layer = layer, stringsAsFactors = FALSE)

#first we need to create a buffer for the MPAs, that approximates the home range size of an otter
buffer_mpas <- function(mpa_layer=cal_mpas, bufferdist=5) {
    #inputs: spatial layer of mpas, buffer distance defined as furthest one-dimensional distance (km) within which an otter would still be considered "associated" or potentially benefitting from the MPA. Somewhat analagous to home range
  
  # spatially buffer the mpa_layer by the given buffer distance, including all records (byid=T)
  buff_lyr <- gBuffer(mpa_layer,width=bufferdist*1000,byid=TRUE)
  
  #clip the buffered layer by the California coastline (so we don't include land area)
  buff_clip <- gDifference(buff_lyr, cal_land,byid=TRUE)
  
  # because the clipped layer loses its attributes, we need to pull out the applicable attributes from the original mpa layer and attach them to the new, clipped layer. We do this by matching row names between the mpa layer and the clip layer
  ids<-sapply(row.names(buff_clip),function(x) strsplit(x, " ")[[1]][1])
  row.names(buff_clip) <- ids
  i<-match(ids,row.names(mpa_layer))
  
  #pull out the appropriate atributes from the original layer...
  attributes <- cal_mpas@data[i,]
  # ...and apply them to our final output
  out.spdf <- SpatialPolygonsDataFrame(buff_clip,attributes)
  plot(out.spdf)
  return(out.spdf)
}
```

So 